hbm_paper_results

### 2024-05-25 paper_lbcc_results.Rmd
cleaning up

df_raw_feat_aov.study %>%
  ggplot() +
  geom_histogram(aes(x=etasq, fill=sig_fdr), alpha=0.5)

df_raw_feat_aov.study %>%
  dplyr::filter(sig_fdr == TRUE) %>%
  nrow()
```
```{r}
raw_centile_df.mean <- raw_centile_df.clean %>%
  pivot_longer(cols=any_of(pheno_list), names_to="pheno", values_to = "centile") %>%
  group_by(participant) %>%
  summarize(study = unique(study),
            site = unique(site),
            age_days=unique(age_days),
            sex=unique(sex),
            mean_cent=mean(centile))
nrow(raw_centile_df.mean) == n_unique(raw_centile_df.mean$participant)
```

@@

raw_centile_df.mean %>%
  ggplot() +
  geom_point(aes(x=age_days/365.25, y=mean_cent, color=site), alpha=0.5) +
  geom_smooth(aes(x=age_days/365.25, y=mean_cent), method="lm") +
  theme(legend.position = "none") +
  facet_wrap(~study)
```
linear model

```{r}
summary(lm(mean_cent ~ study, data=raw_centile_df.mean))
summary(aov(mean_cent ~ study, data=raw_centile_df.mean))
```

@@

df_raw_feat_aov.z.study %>%
  dplyr::filter(sig_fdr == TRUE) %>%
  nrow()
```

@@

```{r}
raw_z_df.mean %>%
  ggplot() +
  geom_point(aes(x=age_days/365.25, y=mean_z), alpha=0.5) +
  geom_smooth(aes(x=age_days/365.25, y=mean_z), method="lm") 
# +
#   facet_wrap(~study)
```

linear model

```{r}
summary(lm(mean_z ~ study, data=raw_z_df.mean))
summary(aov(mean_z ~ study, data=raw_z_df.mean))
```

@@

df_gam_feat_aov.study %>%
  ggplot() +
  geom_histogram(aes(x=etasq, fill=sig_fdr), alpha=0.5)

df_gam_feat_aov.study %>%
  dplyr::filter(sig_fdr == TRUE) %>%
  nrow()
```

@@

linear model

```{r}
summary(lm(mean_cent ~ study, data=gam_centile_df.mean))
summary(aov(mean_cent ~ study, data=gam_centile_df.mean))
```

@@

df_gam_feat_aov.z.study %>%
  dplyr::filter(sig_fdr == TRUE) %>%
  nrow()
```

@@

```{r}
gam_z_df.mean %>%
  ggplot() +
  geom_point(aes(x=age_days/365.25, y=mean_z), alpha=0.5) +
  geom_smooth(aes(x=age_days/365.25, y=mean_z), method="lm") 
# +
#   facet_wrap(~study)
```

linear model

```{r}
summary(lm(mean_z ~ study, data=gam_z_df.mean))
summary(aov(mean_z ~ study, data=gam_z_df.mean))
```

@@

```{r}
ggplot(gamlss_centile_df.clean) +
  geom_point(aes(x=age_days, y=rh_Vol_insula, color=study), alpha=0.5) +
  theme(legend.position="none")
```

@@

df_gamlss_feat_aov.study %>%
  ggplot() +
  geom_histogram(aes(x=etasq, fill=sig_fdr))

df_gamlss_feat_aov.study %>%
  dplyr::filter(sig_fdr==TRUE) %>%
  nrow()
```

@@

```{r}
ggplot(gamlss_centile_df.mean) +
  geom_point(aes(x=(age_days/365.25), y=mean_cent, color=study), alpha=0.5) +
  theme(legend.position="none")

ggplot(gamlss_centile_df.clean) +
  geom_point(aes(x=age_days, y=lh_Vol_inferiortemporal, color=study), alpha=0.2) +
  theme(legend.position="none")
```

```{r}
gamlss_centile_df.mean %>%
  ggplot() +
  geom_boxplot(aes(x=reorder(study, age_days), y=mean_cent)) + 
  #geom_boxplot(aes(x=study, y=mean_cent)) +
  theme(axis.text.x = element_text(angle=90, hjust=0.5))

gamlss_centile_df.mean %>%
  ggplot() +
  geom_point(aes(x=age_days/365.25, y=mean_cent, color=study), alpha=0.5) +
  geom_smooth(aes(x=age_days/365.25, y=mean_cent, color=study), method="lm") +
  theme(legend.position = "none")

gamlss_centile_df.mean %>%
  dplyr::filter(age_days>365) %>%
  ggplot(aes(x=age_days/365.25, y=mean_cent)) +
  geom_point(aes(color=study), alpha=0.7) +
  geom_smooth(method="lm") +
  theme(legend.position="none")
```

linear model

```{r}
summary(lm(mean_cent ~ study, data=gamlss_centile_df.mean))
summary(aov(mean_cent ~ study, data=gamlss_centile_df.mean))
```

@@

df_gamlss_feat_aov.z.study %>%
  dplyr::filter(sig_fdr == TRUE) %>%
  nrow()
```

@@

linear model

```{r}
summary(lm(mean_z ~ study, data=gamlss_z_df.mean))
summary(aov(mean_z ~ study, data=gamlss_z_df.mean))
```

@@

```{r}
summary(lm(mean_cent ~ study, data=gamlss_centile_df.mean))
summary(lm(mean_cent ~ study, data=gam_centile_df.mean))
summary(lm(mean_cent ~ study, data=raw_centile_df.mean))
```

compare magnitude of effect size for study (w/in feature anovas) across unharmonized, combat-gam and combatLS centiles

```{r}
df_gamlss_feat_aov.study$combat <- "gamlss"
df_gam_feat_aov.study$combat <- "gam"
df_raw_feat_aov.study$combat <- "raw"

#put all data frames into list
df_list.tmp <- list(df_gamlss_feat_aov.study, df_gam_feat_aov.study, df_raw_feat_aov.study)

#merge all data frames in list
feat_aov.df <- df_list.tmp %>% reduce(full_join)

feat_aov.df <- feat_aov.df %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

feat_aov.df %>%
  group_by(combat) %>%
  summarize(n_sig = sum(sig_fdr))

feat_aov.df %>%
  ggplot() +
  geom_point(aes(x=etasq, y=pheno, color=combat, shape=sig_fdr), alpha=0.6) +
  facet_wrap(~pheno_cat, scales="free")

feat_aov.df %>%
  dplyr::filter(combat!="raw") %>%
  ggplot() +
  geom_point(aes(x=etasq, y=pheno, color=combat, shape=sig_fdr), alpha=0.6) +
  facet_wrap(~pheno_cat, scales="free")
```
```{r}
gamlss_centile_df.mean$combat <- "gamlss"
gam_centile_df.mean$combat <- "gam"
raw_centile_df.mean$combat <- "raw"

@@

#Feature Centile ~ Mean Study Age

Get study info
```{r}
study_info <- raw_centile_df.clean %>%
  pivot_longer(cols=any_of(pheno_list), names_to="pheno", values_to = "centile") %>%
  group_by(study) %>%
  summarize(study_mean_age=mean(age_years),
            study_median_age=median(age_years),
            study_n= n()) %>%
  dplyr::select(study, study_mean_age, study_median_age, study_n)
```

W/in feat lms

##Unharmonized

```{r}
raw_centile_df.clean_study <- full_join(raw_centile_df.clean, study_info)

df_raw_feat_lm <- data.frame()
for(pheno in raw.feat_to_test){
  # print(pheno)
  df <- tidy(aov(raw_centile_df.clean_study[[pheno]] ~ raw_centile_df.clean_study[["study_mean_age"]]))
  df$pheno <- pheno
  df_raw_feat_lm <- rbind(df_raw_feat_lm, df)
}

df_raw_feat_lm <- df_raw_feat_lm %>%
  dplyr::filter(term != "Residuals") %>%
  dplyr::mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = length(raw.feat_to_test)), #apply fdr correction across all phenos tested
                  sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                                      p.val_fdr >= 0.05 ~ FALSE))
df_raw_feat_lm %>%
  ggplot() +
  geom_bar(aes(y=sig_fdr))

df_raw_feat_lm %>%
  ggplot() +
  geom_histogram(aes(x=statistic))
```

##ComBat-GAM

```{r}
gam_centile_df.clean_study <- full_join(gam_centile_df.clean, study_info)

df_gam_feat_lm <- data.frame()
for(pheno in gam.feat_to_test){
  # print(pheno)
  df <- tidy(aov(gam_centile_df.clean_study[[pheno]] ~ gam_centile_df.clean_study[["study_mean_age"]]))
  df$pheno <- pheno
  df_gam_feat_lm <- rbind(df_gam_feat_lm, df)
}

df_gam_feat_lm <- df_gam_feat_lm %>%
  dplyr::filter(term != "Residuals") %>%
  dplyr::mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = length(gam.feat_to_test)), #apply fdr correction across all phenos tested
                  sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                                      p.val_fdr >= 0.05 ~ FALSE))

df_gam_feat_lm %>%
  ggplot() +
  geom_bar(aes(y=sig_fdr))

df_gam_feat_lm %>%
  ggplot() +
  geom_histogram(aes(x=statistic))
```

```{r}
gamlss_centile_df.clean_study <- full_join(gamlss_centile_df.clean, study_info)

df_gamlss_feat_lm <- data.frame()
for(pheno in gamlss.feat_to_test){
  # print(pheno)
  df <- tidy(aov(gamlss_centile_df.clean_study[[pheno]] ~ gamlss_centile_df.clean_study[["study_mean_age"]]))
  df$pheno <- pheno
  df_gamlss_feat_lm <- rbind(df_gamlss_feat_lm, df)
}

df_gamlss_feat_lm <- df_gamlss_feat_lm %>%
  dplyr::filter(term != "Residuals") %>%
  dplyr::mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = length(gamlss.feat_to_test)), #apply fdr correction across all phenos tested
                  sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                                      p.val_fdr >= 0.05 ~ FALSE))
df_gamlss_feat_lm %>%
  ggplot() +
  geom_bar(aes(y=sig_fdr))

df_gamlss_feat_lm %>%
  ggplot() +
  geom_histogram(aes(x=statistic))
```

*statistic ~ effect size?

```{r}
df_raw_feat_lm$combat <- "raw"
df_gam_feat_lm$combat <- "gam"
df_gamlss_feat_lm$combat <- "gamlss"

#put all data frames into list
df_list.tmp <- list(df_raw_feat_lm, df_gam_feat_lm, df_gamlss_feat_lm)

#merge all data frames in list
all_feat_lm <- df_list.tmp %>% reduce(full_join)

all_feat_lm %>%
  ggplot() +
  geom_histogram(aes(x=statistic, fill=combat, color=combat), position="identity", alpha=0.5, bins=50)
```

#Subject-Level Mean Centiles ~ Mean Study Age

Plots

```{r}
gamlss_centile_df.mean2$combat <- "gamlss"
gam_centile_df.mean2$combat <- "gam"
raw_centile_df.mean2$combat <- "raw"

#put all data frames into list
df_list.tmp <- list(gamlss_centile_df.mean2, gam_centile_df.mean2, raw_centile_df.mean2)

#merge all data frames in list
all_subj_means.df <- df_list.tmp %>% reduce(full_join)

all_subj_means.df %>%
  ggplot(aes(x=study_mean_age, y=mean_cent, color=combat)) +
  geom_point(alpha=0.6) +
  geom_smooth(method="lm")

all_subj_means.df %>%
  ggplot(aes(x=study_mean_age, y=mean_cent, color=combat)) +
  geom_point(alpha=0.3) +
  geom_smooth(method="lm") +
  facet_wrap(~combat)

summary(lm(mean_cent ~ study_mean_age, data=raw_centile_df.mean2))
summary(lm(mean_cent ~ study_mean_age, data=gam_centile_df.mean2))
summary(lm(mean_cent ~ study_mean_age, data=gamlss_centile_df.mean2))
```

```{r}
summary(lm(mean_cent ~ study_n, data=raw_centile_df.mean2))
summary(lm(mean_cent ~ study_n, data=gam_centile_df.mean2))
summary(lm(mean_cent ~ study_n, data=gamlss_centile_df.mean2))

all_subj_means.df %>%
  ggplot(aes(x=study_n, y=mean_cent, color=combat)) +
  geom_point(alpha=0.6) +
  geom_smooth(method="lm")
```
all assoc with study n

@@

@@

mean_cent_deltas_df_study %>%
  ggplot(aes(x=study_mean_age, y=abs_mean_delta_cent)) +
  geom_point(aes(color=study), alpha=0.2) +
  geom_abline(slope = coef(abs_lm)[["study_mean_age"]], 
              intercept = coef(abs_lm)[["(Intercept)"]])+
  theme(legend.position="none")

mean_cent_deltas_df_study <- mean_cent_deltas_df_study %>%
  mutate(age_diff = (age_days/365.25) - study_mean_age,
         abs_age_diff = abs(age_diff))

abs_lm2 <- lm(abs_mean_delta_cent ~ study_mean_age + study_n + abs_age_diff, data=mean_cent_deltas_df_study)
summary(abs_lm2)

@@

@@

```{r}
# study demographics
fig5 <- ggarrange(
ggarrange(sample_boxplot, sample_barplot,
          full_boxplot, full_barplot,
          nrow=2, ncol=2, 
          #align = "h", 
          common.legend = TRUE,
          legend = "bottom",
          widths=c(1.7, 1),
          heights=c(10, 1)),
# boxplots
mean_cents_bxplt,
common.legend=FALSE,
nrow=1, ncol=2,
labels="AUTO",
widths=c(1, 4)
)

ggsave("figs/figure_5_horizontal.jpeg", fig5, width=18, height=9, units="in")
```
alternative vertical stacking

@@

mean_delta_fig <- mean_cent_deltas_df_study %>%
  ggplot(aes(x=study_mean_age, y=abs_mean_delta_cent)) +
  geom_point(aes(color=study), alpha=0.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper,
                    x = study_mean_age), data = pred_df,
                inherit.aes = FALSE, alpha = 0.2) +
    geom_line(data = pred_df, aes(y = fit, x=study_mean_age))+
  #theme(legend.position="none") +
  labs(x="Mean Age of Primary Study Sample (Years)", y="Mean Absolute Difference in Centile Scores (Centiles)", title="Subject-wise Mean Distance Between Centiles Calculated From \nComBatLS and ComBat-GAM-harmonized Data")
ggsave("figs/supp_mean_delta_fig.jpeg", mean_delta_fig, width=10, height=7, units="in")

mean_cent_deltas_df_study %>%
  dplyr::filter(abs_mean_delta_cent > 4) %>%
  dplyr::select(study, study_mean_age)

mean_cent_deltas_df_study%>%
  dplyr::filter(study == "Cornell_C2") %>%
  nrow()
```

Cornell_C2 has very low n=5 -> big differences in estimation

@@

mean_cent_deltas_df_study %>%
  ggplot(aes(x=abs_age_diff, y=abs_mean_delta_cent)) +
  geom_point(alpha=0.05, aes(color=study)) +
  geom_ribbon(aes(ymin = lower, ymax = upper,
                    x = abs_age_diff), data = pdat2,
                inherit.aes = FALSE, alpha = 0.2) +
    geom_line(data = pdat2, aes(y = fit, x=abs_age_diff)) +
  theme(legend.position="none") +
  labs(x="Subject's Offset from Mean Age of Primary Study (Years)", y="Mean Absolute Difference in Centile Scores (Centiles)", title="Subject-wise Mean Distance Between Centiles Calculated From \nComBatLS and ComBat-GAM-harmonized Data")
```

>>>>

table(df_raw_feat_aov.study$sig_fdr)
```
```{r}
raw_centile_df.mean <- raw_centile_df.clean %>%
  pivot_longer(cols=any_of(pheno_list), names_to="pheno", values_to = "centile") %>%
  group_by(participant) %>%
  summarize(study = unique(study),
            site = unique(site),
            age_days=unique(age_days),
            sex=unique(sex),
            mean_cent=mean(centile))
nrow(raw_centile_df.mean) == n_unique(raw_centile_df.mean$participant)
```

@@

raw_centile_df.mean %>%
  ggplot() +
  geom_point(aes(x=age_days/365.25, y=mean_cent, color=site), alpha=0.5) +
  geom_smooth(aes(x=age_days/365.25, y=mean_cent), method="lm") +
  theme(legend.position = "none") +
  facet_wrap(~study)
```

ANOVA of subjects' mean centiles across studies

```{r}
summary(aov(mean_cent ~ study, data=raw_centile_df.mean))
```

@@

table(df_raw_feat_aov.z.study$sig_fdr)
```

@@

ANOVA of subject means

```{r}
summary(aov(mean_z ~ study, data=raw_z_df.mean))
```

@@

table(df_gam_feat_aov.study$sig_fdr)
```

@@

ANOVA

```{r}
summary(aov(mean_cent ~ study, data=gam_centile_df.mean))
```

@@

table(df_gam_feat_aov.z.study$sig_fdr)
```

@@

```{r}
gam_z_df.mean %>%
  ggplot() +
  geom_point(aes(x=age_days/365.25, y=mean_z), alpha=0.5) +
  geom_smooth(aes(x=age_days/365.25, y=mean_z), method="lm") 
```

Subject mean ANOVA

```{r}
summary(aov(mean_z ~ study, data=gam_z_df.mean))
```

@@

@@

table(df_gamlss_feat_aov.study$sig_fdr)
```

@@

```{r}
gamlss_centile_df.mean %>%
  ggplot() +
  geom_boxplot(aes(x=reorder(study, age_days), y=mean_cent)) + 
  #geom_boxplot(aes(x=study, y=mean_cent)) +
  theme(axis.text.x = element_text(angle=90, hjust=0.5))
```

subject mean ANOVA

```{r}
summary(aov(mean_cent ~ study, data=gamlss_centile_df.mean))
```

@@

table(df_gamlss_feat_aov.z.study$sig_fdr)
```

@@

subject mean ANOVA

```{r}
summary(aov(mean_z ~ study, data=gamlss_z_df.mean))
```

@@

```{r}
gamlss_centile_df.mean$combat <- "gamlss"
gam_centile_df.mean$combat <- "gam"
raw_centile_df.mean$combat <- "raw"

@@

Cohen's F squared for study effects

```{r}
gam_centile_f2 <- fread(file=paste0(der_path, "/lifespan-CN-imp-sites-euler-log-cf-cf.gam-batch.study-data_cohenfsq.csv"))
gam_centile_f2$combat <- "ComBat-GAM"

gamlss_centile_f2 <- fread(file=paste0(der_path, "/lifespan-CN-imp-sites-euler-log-cf-cf.gamlss-batch.study-data_cohenfsq.csv"))
gamlss_centile_f2$combat <- "ComBatLS"

f2.df <- rbind(gam_centile_f2, gamlss_centile_f2)
f2.df <- f2.df %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

fsq.df <- ggplot(f2.df) +
  geom_col(aes(x=pheno, y=fsq, fill=combat), position="dodge") + 
  facet_wrap(~pheno_cat, scales="free") +
  theme_linedraw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  labs(x = "Feature", y = "Effect Size (Cohen's F-Squared)", fill="ComBat")

ggsave("figs/supp_fsq.jpeg", fsq.df, width=8, height=6, units="in")
```

```{r}
f2.df.sum <- f2.df %>%
  group_by(pheno) %>%
  arrange(combat) %>%
  mutate(F2_diff = fsq - fsq[1]) #gamlss - gam value, + number means site effect is larger in combatLS

ggplot(f2.df.sum) +
  geom_col(aes(x=pheno, y=F2_diff), position="dodge") + 
  facet_wrap(~pheno_cat, scales="free") +
  theme(axis.text.x = element_blank())

f2.df.sum %>%
  group_by(pheno_cat) %>%
  summarise(mean_fsq_diff = mean(F2_diff)) %>%
  ggplot() + 
  geom_col(aes(x=pheno_cat, y=mean_fsq_diff), position="dodge")
```

@@

Get study info
```{r}
nrow(raw_centile_df.clean) == n_unique(raw_centile_df.clean$participant)

study_info <- raw_centile_df.clean %>%
  group_by(study) %>%
  summarize(study_mean_age=mean(age_years),
            study_median_age=median(age_years),
            study_n = n()) %>%
  dplyr::select(study, study_mean_age, study_median_age, study_n)
```

@@

mean_cent_deltas_df_study %>%
  ggplot(aes(x=study_mean_age, y=abs_mean_delta_cent)) +
  geom_point(aes(color=study), alpha=0.2) +
  geom_abline(slope = coef(abs_lm)[["study_mean_age"]], 
              intercept = coef(abs_lm)[["(Intercept)"]])+
  theme(legend.position="none")
```

See if a subject's distance from their study's mean age is assoc with how much their centile changes

```{r}
mean_cent_deltas_df_study <- mean_cent_deltas_df_study %>%
  mutate(age_diff = (age_days/365.25) - study_mean_age,
         abs_age_diff = abs(age_diff))

abs_lm2 <- lm(abs_mean_delta_cent ~ study_n + abs_age_diff, data=mean_cent_deltas_df_study)
summary(abs_lm2)

@@

## Subject-Level Mean Centiles ~ Mean Study Age

Make sure that results of mean centile delta ~ mean study age are related to the difference itself, not because ComBatLS and/or ComBat-GAM centiles are associated with mean study age.

NS

NS

NS

@@

Reformatting plots for prettier composite fig

@@

mean_delta_fig <- mean_cent_deltas_df_study %>%
  ggplot(aes(x=study_mean_age, y=abs_mean_delta_cent)) +
  geom_point(aes(color=study), alpha=0.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper,
                    x = study_mean_age), data = pred_df,
                inherit.aes = FALSE, alpha = 0.2) +
    geom_line(data = pred_df, aes(y = fit, x=study_mean_age))+
  theme_linedraw() +
  theme(legend.position="none") +
  labs(x="Mean Age of Primary Study Sample (Years)", y="Mean Absolute Difference in Centile Scores (Centiles)", title="Subject-wise Mean Distance Between Centiles Calculated From ComBatLS and ComBat-GAM \n-harmonized Data")
ggsave("figs/supp_mean_delta_fig.jpeg", mean_delta_fig, width=9, height=6, units="in")
```

lines show marginal effects of mean study age

@@

mean_delta_fig2 <- mean_cent_deltas_df_study %>%
  ggplot(aes(x=abs_age_diff, y=abs_mean_delta_cent)) +
  geom_point(alpha=0.05, aes(color=study)) +
  geom_ribbon(aes(ymin = lower, ymax = upper,
                    x = abs_age_diff), data = pdat2,
                inherit.aes = FALSE, alpha = 0.2) +
    geom_line(data = pdat2, aes(y = fit, x=abs_age_diff)) +
  theme_linedraw() +
  theme(legend.position="none") +
  labs(x="Subject's Offset from Mean Age of Primary Study (Years)", y="Mean Absolute Difference in Centile Scores (Centiles)", title="Subject-wise Mean Distance Between Centiles Calculated From ComBatLS and ComBat-GAM \n-harmonized Data")
ggsave("figs/supp_mean_delta_fig2.jpeg", mean_delta_fig2, width=9, height=6, units="in")
```
